---
title: "Text hex logos with {hext}"
date: 2026-02-15
slug: "turn-tables"
categories:
  - hext
  - r
---

```
  ________
 /        \
/   he     \
\     xt   /
 \________/
```

## tl;dr

The [{hext} R package](https://github.com/matt-dray/hext) helps you make text-based hex logos.

## A simple mind

Call me a normie, but I am an R-hex-logo enjoyer. 

I took a step further last year to create my own (limited) hex-making package, [{gex}](https://github.com/matt-dray/gex), which I've [written about](https://www.rostrum.blog/posts/2025-02-25-gex/) before[^craig].

The goal of {gex} was to keep things simple: minimal dependencies, only grid graphics[^base].

But I was too short-sighted, too small-brained.
I went simple, but _not simple enough_.

## Do the simple things well

Yihui recently said '[bye, hex stickers](https://yihui.org/en/2026/02/bye-stickers/)'.

In his post he discussed an entirely text-based hex logo for [{litedown}](https://github.com/yihui/litedown/blob/main/README.md). Charlie Gao's [{secretbase}](https://github.com/shikokuchuo/secretbase/blob/main/README.md)[^sb] has also joined the text revolution.

The spirit of these packages is simplicity. 
Their logos match this aesthetic.

Simplicity is a probably a good goal in general.

So, naturally, I have overengineered a simple thing to make it simple for you to not think about the complication behind such a simple thing.

## Next: text hext

[{hext}](https://github.com/matt-dray/hext) lets you write text-based logos.

You could install like:

```{r}
#| label: install-hext
#| eval: false
pak::pak("matt-dray/hext")  # GitHub-only
```

There is only one function: `hext()`.
(This is a feature, not a bug.)

Hexes have four lines[^length] to which you can add text and specify alignment.

```{r}
#| label: hext-example
hext::hext(
  "keep", "it", "simple,", "silly",
  "left", "centre", "right", "left"
)
```

`hext()` handles character padding and line-length limits on your behalf.
It tries its best with Unicode (like emoji)[^render], but rendering is hard.

That's it, really.

## Hexagone

Think less, say less.

### Environment {.appendix}

<details><summary>Session info</summary>
```{r sessioninfo, eval=TRUE, echo=FALSE}
cat("Last rendered:", format(Sys.time(), usetz = TRUE)); sessionInfo()
```
</details>

[^craig]: You know who you are, dear one-user-who-isn't-me
[^sb]: See-also the namesake YouTube channel for sports videos. And documentaries about the history of slipping on banana peels.
[^base]: The irony being that the original iteration—called {hexbase}—was even simpler. Dependency-free, in fact.
[^length]: The community seems divided on the number of characters per row. Like Charlie, I have opted for 8-10-10-8 top-to-bottom, which gives a little extra character space, even if makes the hexagon look a little more plump.
[^render]: This works via `nchar(x, type = "width")` (or `stringri::stri_width()` if you have {stringi} installed), rather than the default `nchar(x, type = "chars")`. The smiley emoji has a width of `2`, but is `1` character long, for example. Obviously this impacts the whitespace padding that `hext()` needs to apply. If this is bothersome, you can use the argument `count_type` in `hext()` to switch between these counting modes.